---
title: "Bootstrap"
author: "Daniel Del Rosso"
format: html
editor: visual
embed-resources: true
---

We first load our packages and data:

```{r}
#| include = FALSE
library(boot)
library(here)
library(dplyr)

finaldata <- read.csv(here("output_files/finaldata.csv"))
```

We then filter by year 2017:

```{r}
finaldata2017 <- finaldata %>% 
  filter(year == 2017) 
```

We define the function for the boot package to use:

```{r}
getmeddiff <- function(data, indices, variable) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data[[variable]], sample_data$conflict, FUN = function(x) median(x))
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
}

```

We do some final filtering and run the bootstrap. We subsequently calculate the confidence intervals.

```{r}
finaldata2017_neonatal <- finaldata2017 %>% 
  filter(!is.na(neonatal))

finaldata2017_under5 <- finaldata2017 %>% 
  filter(!is.na(under5))


finaldata2017_infant <- finaldata2017 %>% 
  filter(!is.na(infant))

bootout_neonatal <- boot(finaldata2017_neonatal, statistic = getmeddiff, strata = finaldata2017_neonatal$conflict, variable = "neonatal",  R = 1000)

bootout_under5 <- boot(finaldata2017_under5, statistic = getmeddiff, strata = finaldata2017_under5$conflict, variable = "under5", R = 1000)

bootout_infant <- boot(finaldata2017_infant, statistic = getmeddiff, strata = finaldata2017_infant$conflict, variable = "infant", R = 1000)

```

```{r}
boot.ci(boot.out = bootout_neonatal, conf = 0.95, type = c("basic", "perc", "bca"))

```

```{r}
boot.ci(boot.out = bootout_under5, conf = 0.95, type = c("basic", "perc", "bca"))

```

```{r}
boot.ci(boot.out = bootout_infant, conf = 0.95, type = c("basic", "perc", "bca"))

```

For infant mortality, neonatal mortality, and under-5 mortality, we have significacnt evidence that the difference in medians between countries experiencing conflict and non-conflict is greater than 0, at $\alpha = 0.05$.
